---
title: "Examples of the use of Sapply"
output: html_notebook
---

The sapply() function applies a function to individual values of a dataframe, and simplifies the output.

Structure of the sapply() function: sapply(data, function, ...)

The data used in this example is

```{r}

dataset1 <- data.frame(observation_A = 16:8, observation_B = c(20:19, 6:12))

dataset1
```

###Exercise 1###

Using sapply(), find the length of dataset1‘s observations:

```{r}
sapply(dataset1, length)

```


###Exercise 2###

Using sapply(), find the sums of dataset1‘s observations:

```{r}
sapply(dataset1, sum)
```


###Exercise 3###

Use sapply() to find the quantiles of dataset1‘s columns:

```{r}
sapply(dataset1, quantile)
```


###Exercise 4###

Find the classes of dataset1‘s columns:

```{r}
sapply(dataset1, class)

```


###Exercise 5###

Required function:

```{r}
DerivativeFunction <- function(x) { log10(x) + 1 }
```

Apply the “DerivativeFunction” to dataset1, with simplified output:

```{r}
sapply(dataset1, DerivativeFunction)
```


###Exercise 6###

Script the “DerivativeFunction” within sapply(). The data is dataset1:

```{r}
sapply(dataset1, function(x) log10(x) + 1)
```


###Exercise 7###

Find the range of dataset1:

```{r}
sapply(dataset1, range)
```


###Exercise 8###

Print dataset1 with the sapply() function:

```{r}
sapply(dataset1, print)
```


###Exercise 9###

Find the mean of dataset1‘s observations:

```{r}
sapply(dataset1, mean)
```


###Exercise 10###

Use sapply() to inspect dataset1 for numeric values:

```{r}
sapply(dataset1, is.numeric)
```


